An open and continuously updated fern tree of life (FTOL)

### 2022-01-30 ms/manuscript.Rmd
Update MS

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
```

```{r load-targets}
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_tree,
		plastome_metadata_renamed,
		ppgi_taxonomy),
	store = here::here("ftol_cache")
)
```

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< 7000 bp) or plastome sequences (> 7000 bp) with the "[SLEN]" term.
Furthermore, there is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(glue)
library(phytools)

# Specify percentage format
percent <- function(...) {scales::percent(...)}
number <- function(...) {scales::number(big.mark = ",", ...)}

# Specify FTOL version
ftol_ver <- "v0.0.2"
```

```{r load-targets}
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_tree,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat),
	store = here::here("ftol_cache")
)
```

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< 7000 bp) or plastome sequences (> 7000 bp) with the "[SLEN]" term.
Furthermore, there is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.

@@

## Molecular dating

We dated the tree using treePL [@Smith2012], which is one of the only programs capable of dating very large alignments.
We applied a set of `r nrow (plastid_calibration_dates)` calibration points, mostly based on @Testo2016a and references within.

@@

# Results

## Sampling

```{r sampling-stats, message=FALSE}
# Make list of FTOL sampling
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

# Make list of all accepted species
accepted_species <-
	pteridocat %>% 
	# Only keep accepted names
	filter(str_detect(taxonomicStatus, "accepted")) %>%
	# Only keep species
	# FIXME: need to add rank for newly added names in pteridocat_maker
	# then don't need is.na()
	filter(taxonRank == "species" | is.na(taxonRank)) %>%
	select(scientificName) %>%
	mutate(
		rgnparser::gn_parse_tidy(scientificName) %>% select(species = canonicalsimple),
		species = str_replace_all(species, " ", "_")
	) %>%
	# Map on higher level taxonomy
	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
	left_join(select(ppgi_taxonomy, genus, family, order, class), by = "genus") %>%
	# FIXME: update class for nothogenera in PPG
	mutate(class = case_when(
		str_detect(scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
		TRUE ~ class
	)) %>%
	# Drop lycophytes
	assert(not_na, class) %>%
	filter(class == "Polypodiopsida") %>%
	select(-class)

accepted_sampling_by_rank <-
	accepted_species %>%
	select(-scientificName) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup %>%
	# Note that nothogenera have no family or order
	filter(!is.na(name))

coverage <- left_join(
	ftol_sampling_by_rank,
	rename(accepted_sampling_by_rank, n_accepted = n),
	by = c("rank", "name")
) %>%
	mutate(coverage = n/n_accepted)

# Format overall sampling for printing in MS
total_sampling <-
	# Start with FTOL sampling by genus, family, order
	ftol_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		sanger_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted),
		coverage = glue("{number(n)}/{number(n_accepted)}")) %>%
	# Split for printing
	split(.$rank)
```

Taxon sampling for the current version of FTOL (`r ftol_ver`) includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).

# Figures

```{r backbone-tree}
og_taxa <-
	sanger_sampling %>%
	filter(outgroup == TRUE) %>%
	pull(species)

backbone_tree_full <- 
	plastome_tree %>%
	ape::root.phylo("Anthoceros_angustus") %>%
	ape::ladderize() %>%
	ape::drop.tip(og_taxa)

family_exemp <-
tibble(
	species = backbone_tree_full$tip.label
) %>%
	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
	group_by(family) %>%
	slice(1) %>%
	ungroup() %>%
	pull(species)

backbone_tree_exemp <- ape::keep.tip(backbone_tree_full, family_exemp)

bb_trans_table <-
	tibble(species = backbone_tree_exemp$tip.label) %>%
	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
	left_join(accepted_sampling_by_rank, by = c(family = "name")) %>%
	assert(not_na, species, family, n) %>%
	select(tip.label = species, clade.label = family, N = n) %>%
	## set crown node depth to 1/2 the maximum depth
	mutate(
		depth = sapply(backbone_tree_exemp$tip.label, function(x,y) 
	0.5*y$edge.length[which(backbone_tree_exemp$edge[,2] ==
	which(y$tip.label == x))], y = backbone_tree_exemp)
	)

## convert
tt <- phylo.toBackbone(backbone_tree_exemp, bb_trans_table)
## plot
plot(tt)
```
